---
title: "Spatial Analysis"
---


```{r, include=FALSE}
library(sf)
wake_r1=read.csv("projects/data/WakeCrime.csv", nrows = 1)
```

Here I investigate crime data in the Neuse township of Wake County, NC from 2015 to 2026. This data comes from Wake County's Public [Database](https://data.wake.gov/search){target="_blank" rel="noopener"} and contains all police reports from the Raleigh Police Department. To analyze this data, I will be running this data in R; below is a row from the dataset.  

```{r, echo=FALSE}
wake_r1
```

As we can see, there are very specific details for each row or report. I am only going to be looking at traffic crimes that occurred in the Neuse township.  
Township geometry data is also available from the database mentioned earlier, below we can see how that data is visualized.  
```{r}
shapeWake <- st_read("projects/data/Townships-shp/townships.shp", quiet=T)
shapeWake=st_transform(shapeWake, "EPSG:4326")
plot(shapeWake["NAME"], main=NA, key.length = 1)
```
I am interested in viewing the difference, or lack thereof, between the occurrence of traffic crimes on weekdays and weekends. Below we filter and partition the data for this investigation.  
```{r, eval=FALSE}
weekends <- c("Saturday", "Sunday")
# respective data sets for weekdays and weekends
wake1 <- wake[wake$Reported.Day.of.Week %in% weekends &
                wake$Crime_Category=="TRAFFIC",]
wake2 <- wake[!(wake$Reported.Day.of.Week %in% weekends) &
                wake$Crime_Category=="TRAFFIC",]
```

Next, we can view the collective point pattern of traffic crimes in Neuse.  
![Point pattern of traffic crimes in Neuse, 2015-2026](projects/results/sa_p1){width=700}

Is there a difference in the point patterns between traffic crimes on weekdays and on weekends? By comparing these two point patterns, we will be able to see if and where there are regions of concentrated traffic crimes partitioned into weekdays and weekends.  
First to verify the validity of this question, I will be testing to see if the point patterns for these data show complete spatial randomness. Then I will be comparing the intensities of each point pattern.  

### K-function Tests for Complete Spatial Randomness  
To test for complete spatial randomness, I am going to compute the K function for both point patterns.  
Assuming complete spatial randomness, the K function should be approximately equal or similar to the curve represented by the function $\pi h^2$, while a greater value of K implies positive clustering in the point patterns.  
The K-function curves for both point patterns are plotted below. A spatial lag of 0.01 represents approximately one kilometer. The K-functions are computed over a lag of about 3km.  

The curves for both groups strongly suggest positive clustering in the point patterns. These results aren't very surprising if we consider that traffic crimes are confined to roads rather than any location and are still susceptible to external factors such as areas with heavy traffic.  
Using kernel smoothing, we can also plot the intensity of both patterns.  

### Comparing Weekend and Weekday Point Pattern Intensity  
We can compare the intensities between these point patterns to see if where the point patterns differ. We can accomplish this using the log relative risk, similar to a hazard ratio where we want to obtain the ratio of the normalized intensities. Assuming that the proportions of weekend and weekday traffic crimes are equal to one, our estimate for the log relative risk (r(s)) is given by:  
$r(s)=ln(\frac{\lambda_1(s)}{\lambda_0(s)})-ln(\frac{N_1}{N_0})$  
Here, $\lambda_1$ and $\lambda_0$ are the intensities for the weekend and weekday point patterns respectively. $N_1$ and $N_0$ are the sample sizes for each point pattern. Computing this function over all of the spatial domain $s$, we obtain the map of log relative risk:  

We can see some areas do not have a compute relative risk which is due to a lack of observed points in those regions. For areas with observed risk, most regions seem close to zero, making it difficult to determine which regions have significantly large or low values of log relative risk.  
By quantifying the variability of the log relative risk using Monte Carlo simulations, we can use tolerance intervals to identify where the risk is significantly different from zero.  
To obtain the tolerance intervals, I conducted 200 simulations, re-sampling weekend and weekday traffic crimes at each of the observed traffic crime locations before recomputing the log relative risk. The 95% tolerance intervals were obtained using the 97.5th and 2.5th quantiles over each simulation.  
The plot below shows the regions where the log relative risk either fell above or below the Monte Carlo tolerance intervals.  

We can see that traffic crimes seem to increase near Southwest Neuse(roughly around Atlantic Avenue) on weekends. On weekdays, the intensity of traffic crimes seems to increase sparsely on the borders and interestingly enough further up on Atlantic Avenue near Millbrook. The areas in green might be describing areas that are commonly used for work commute on weekdays. It might be helpful to know what the exact traffic crimes committed were for each observation, such as whether these were speeding or reckless driving infractions.  